---
title: "class11"
format: pdf
author: "Kate Zhou (PID:A17373286)"
toc: True
---

Here we read the results from AlphaFold and try to interpret all the models and quality score metrics:

```{r}
library(bio3d)

pth <- "dimmer_23119/"
pdb.files <- list.files(path=pth, full.names = TRUE, pattern = ".pdb")
```

Align and superpose all these models
```{r}
file.exists(pdb.files)
```
```{r}
pdbs <- pdbaln(pdb.files, fit = TRUE, exefile="msa")
```
```{r}
#view.pdbs(pdbs)
```

```{r}
plot(pdbs$b[1,], typ="l", ylim=c(0, 100), ylab="pLDT Score")
lines(pdbs$b[2,], typ="l", col="blue")
lines(pdbs$b[3,], typ="l", col="green")
lines(pdbs$b[4,], typ="l", col="orange")
lines(pdbs$b[5,], typ="l", col="red")
```
```{r}
rd <- rmsd(pdbs, fit=T)

library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

# Predicted Alignment Error for domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=pth,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

# Score Residue Conservation from alignment file

AlphaFold returns its's large alignment file used for analysis. Here we read this file and score conservation per position
```{r}
aln_file <- list.files(path=pth,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

Find the consensus sequence at a very high cut-off to find invariant residues
```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```
