---
title: "class18"
format: pdf
author: "Kate Zhou (PID:A17373286)"
toc: TRUE
---

## Background

Pertussis (a.k.a whooping cough) is a common lung infection caused by the bacteria *B.Pertussis*.

The CDC tracks cases of Pertussis in the US:
https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html

## Examining cases of Pertussis by year

We can use 

```{r, include=FALSE}
cdc <- data.frame(
 year = c(1922L,1923L,1924L,1925L,
          1926L,1927L,1928L,1929L,1930L,1931L,
          1932L,1933L,1934L,1935L,1936L,
          1937L,1938L,1939L,1940L,1941L,1942L,
          1943L,1944L,1945L,1946L,1947L,
          1948L,1949L,1950L,1951L,1952L,
          1953L,1954L,1955L,1956L,1957L,1958L,
          1959L,1960L,1961L,1962L,1963L,
          1964L,1965L,1966L,1967L,1968L,1969L,
          1970L,1971L,1972L,1973L,1974L,
          1975L,1976L,1977L,1978L,1979L,1980L,
          1981L,1982L,1983L,1984L,1985L,
          1986L,1987L,1988L,1989L,1990L,
          1991L,1992L,1993L,1994L,1995L,1996L,
          1997L,1998L,1999L,2000L,2001L,
          2002L,2003L,2004L,2005L,2006L,2007L,
          2008L,2009L,2010L,2011L,2012L,
          2013L,2014L,2015L,2016L,2017L,2018L,
          2019L,2020L,2021L,2022L,2023L,2024L),
cases = c(107473,164191,165418,152003,
          202210,181411,161799,197371,
          166914,172559,215343,179135,265269,
          180518,147237,214652,227319,103188,
          183866,222202,191383,191890,109873,
          133792,109860,156517,74715,69479,
          120718,68687,45030,37129,60886,
          62786,31732,28295,32148,40005,
          14809,11468,17749,17135,13005,6799,
          7717,9718,4810,3285,4249,3036,
          3287,1759,2402,1738,1010,2177,2063,
          1623,1730,1248,1895,2463,2276,
          3589,4195,2823,3450,4157,4570,
          2719,4083,6586,4617,5137,7796,6564,
          7405,7298,7867,7580,9771,11647,
          25827,25616,15632,10454,13278,
          16858,27550,18719,48277,28639,32971,
          20762,17972,18975,15609,18617,
          6124,2116,3044,7063,35493)
       )
```

```{r}
head(cdc)
```

> Q1. Make a plot of pertussis cases peryear using ggplot

```{r}
library(ggplot2)

cases <-  ggplot(cdc) +
          aes(year,cases) +
          geom_point(col = "violet") +
          geom_line() 

cases
```

> Q2. Add some key time points in our history of interaction with Pertussis. These includewP roll-out (the first vaccine) in 1945 and the switch to aP in 1996

We can use `geom_vline()` for this

```{r}
cases + geom_vline(xintercept =1945, col="blue") +
       geom_vline(xintercept =1996, col="red") +
       geom_vline(xintercept = 2020, col="purple")
```
> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Mounting ecidence suggests that the newer **aP** vaccine is less effective over the long term than the older **wP** vaccine that it replaced. In other, words, vaccine protection wanes more rappidly with aP than with wP.

## Enter the CMI-PB project

CMI-PB (computational Models of Immunity - Pertussis boost) major goal is to incestigate how the immune responds differently to with aP vs wP vaccinated individuals and be able to predict this at an erarly stage.

CMI-PB makes all their collected data freely avavilable and they store it in a database composed different tables. 

We can use the **jsonlite** package to read this data
```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

> Q. How many subjects (i.e. enrolled people) are there in this dataset?

```{r}
nrow(subject)
```

> Q. How many "aP" and "wP" subjects are there?

```{r}
table(subject$infancy_vac)
```

> Q. How many Male/Female are in the dataset?

```{r}
table(subject$biological_sex)
```


> Q. How about gender and race number

```{r}
table(subject$race,subject$biological_sex)
```

> Q. Age

```{r}

```


> Q. Is this representative of the US population?

No, UCSD student population

Let's read another database table from CMI-PB

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = TRUE)

ab_data <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", 
                      simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_data)
```

We want to "join" these tables to get all our information together. For this we will use the **dplyr** package and `inner_join()` function.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

One more "join"to get ab_data

```{r}
abdata <- inner_join(meta, ab_data)

head(abdata)
```
```{r}
dim(abdata)
```


> Q. How many Ab isotypes are there in the dataset?

```{r}
table(abdata$isotype)
```

> Q. How many different antigens are measured in the dataset?

```{r}
table(abdata$antigen)
```

> Q. Make a boxplot of antigen levels accross the whole dataset

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```

> Q. Are there obvious differenes between aP and wP values

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot()
```

Or we can `infancy_vac` to get two individual plots one for each value of `infancy_vac`

```{r}
ggplot(abdata) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

viral infections.

```{r}
igg <- abdata |> 
  filter(isotype == "IgG")

head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```
Focus in furthwer in jsut one of these anitgens - let's pick **PT** (Pertussis Toxin, one of the main toxins of the bacteria)

```{r}
table(igg$dataset)
```

```{r}
pt_igg <- abdata |>
          filter(isotype == "IgG",
                 antigen == "PT",
                 dataset == "2021_dataset")

head(pt_igg)
```

```{r}
dim(pt_igg)
```

```{r}
ggplot(pt_igg) +
  aes(actual_day_relative_to_boost, MFI_normalised, 
      col = infancy_vac,
      # connect lines with the same subject_id
      group=subject_id) +
  geom_point() +
  geom_line() +
  theme_bw() +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 14)
```

